Exploring the Vectorization of Python Constructs Using Pythran and Boost SIMD

### 2013-11-29 doc/papers/wpmvp14/paper.tex
+Spellchecking

The Python language~\cite{rossum97} has grown in audience for the past ten
years, even reaching the world of scientific computations~\cite{oliphant2007}
thanks to the \texttt{numpy} module~\cite{numpyarray2011}, a module that
provides an efficient multi-dimensionnal array type, and the \texttt{scipy}
package~\cite{scipy} that provides a MATLAB-like API. As a consequence, more
and more code is being written either in pure Python, generally to prototype an
application, or as a Python and native code mix when performance matters.
However, Python trades performance for dynamicity and does not particularly
shines in terms of performance.

@@

    \item[dynamic binding] each variable access is made through a dictionnary
        look-up to bind variable names to variable instances~;

@@

Single Instruction Multiple Data (SIMD) extensions have been a feature of
choice for processor manufacturers for a couple of decades. However,
programming applications that take advantage of the SIMD extension available on
the current target remains a complex task. Programmers that use low-level
intrinsics have to deal with a verbose programming style due to the fact that
SIMD instructions sets cover a few common functionnalities, requiring to bury
the initial algorithms in architecture specific implementation details.
Furthermore, these efforts have to be repeated for every different extension
that one may want to support, making design and maintenance of such
applications very time consuming.

\boostsimd{} is a high-level C++ library to program SIMD architectures. 
Designed as an \textit{Embedded Domain Specific Language}, \boostsimd{} 
provides both expresiveness and performance by using generic programming 
to handle vectorization. \boostsimd{} does not only provide a portable 
way but also enable the use of common programming idioms when designing 
SIMD-aware algorithms. Two components are provided:

@@

Previous examples of integration within standard algorithm are still limited.
Applying \texttt{transform} or \texttt{fold} algorithm on SIMD aware data
requires the size of the data to be an exact multiple of the SIMD register width
and, in the case of fold, to potentially perform additional operations at the
end of its call. To alleviate this limitation, \boostsimd{}
provides its own overload for both \texttt{transform} and \texttt{fold} that take
care of potential trailing data and performs proper completion of \texttt{fold}.
Moreover, as \boostsimd{} provides default implementation of all its function for scalar values,
we can rely on polymorphic callable object to write a mixed scalar/SIMD call to algortihm
like \texttt{transform}. By using the \texttt{boost::simd::transform} algorithm, code
handlign alignment, scalar prologue and epilogue is generated.

@@

The first attempt to have Pytrhan generate vectorized code directly relied on
SSE intrinsics. This approach quickly proved to be inappropriate for several
reasons:

@@

    \item[completness] Many basic mathematic functions such as cosinus or sinus
        are not directly available in the target instruction sets;

@@

Pythran uses function objects to represent any Python functions, either
user-functions or functions from the standard library. The former are
automatically generated from the user code analysis, and the latter are
hand-coded in a library called \texttt{pythonic++} provided with the compiler.
For instance the Python function \texttt{def foo(x): return 3 * x} is compiled
into a code similar to the one in Listing~\ref{lst:simple-pythran-output}. The
actual would be more compilcated to handle type inference across several
statements, but the structure remains valid.

@@

A high-level language such as Python exhibits another level of vectorization
opportunity in implict loop. Implicit loops can be found in many places, but
this paper focuses on \emph{list comprehension} and \emph{generator
expression}, \emph{map} or \emph{sum} intrinsics and \emph{array operations}
through the \texttt{numpy} module.

@@

The \texttt{sum} intrinsic is a typical case of potential vectorization of an
implict loop. Its prototype is: \texttt{sum(sequence[, start]) $\rightarrow$
value} and its behavior is equivalent to the Python code in
Listing~\ref{lst:sum-py}. Depending on the type of \texttt{sequence}, and also
wether one needs to strictly conforms to the IEEE Standard for Floating-Point
Arithmetic (IEEE 754), vectorization of this reduction is a comm case of \texttt{fold}.

@@

Appart from standard compliancy, to be able to efficiently vectorize this code in C++, one must check that:

@@

Condition~\ref{enu:same-types} is a prerequiste for static typing and efficient
unboxing. For instance, any code that does not match it will fail to compile
using Pythran or Shed-Skin. It is also satisified for the \texttt{ndarray}
container as they wraps unboxed arrays. Condition~\ref{enu:scalar-types} can be
checked at compile time. Python supports two native scalar types candidates for
vectorization: \texttt{int} (64 bits integer) and \texttt{float}
(double-precision floating point numbers). The \texttt{numpy} model add supports for a
wide range of signed and unsigned integers as well as single, double or
quadruple precision float. Depending on the target architecture, only some of
them can fit into vector registers. Finally condition~\ref{enu:contiguous} also
is a static property of a container. It is verified for C++ alternative to
Python's \texttt{list}, \texttt{str} and \texttt{ndarrays}, but generally not
for \texttt{set} or \texttt{dict}. Condition \ref{enu:random-access} is not
matched by \emph{generator}s, the Python incarnation of co-routines.

@@

As function have to be generic but some containers do not match requierement
properties, it is possible to select a vectorized
implementation of fallback to a generic one using Substitution Failure
Is Not An Error (SFINAE)~\cite{metaprogramming2002}.

@@

Python provides \emph{list comprehension} as a way to create \texttt{list} from
any iterable. For instance \texttt{[abs(x) for x in l]} creates a new
\texttt{list} that holds the absolute value of each element of \texttt{l}. A
filter can be set up as in  \texttt{[abs(x) for x in l if x \% 3]}, where only
the elements non multiple of 3 are taken into account. Let us call
\emph{comprehesion rule} the expression on the left of the
\texttt{for}.

List comprehension can be translated into a combination of \texttt{map} and
(optionnaly) \texttt{filter}, by transformation of the comprehension rule into
a lambda function. For instance the previous example can be re-written as
\texttt{map(abs, filter(lambda x: x\% 3, l))}. Once in this form, the same
    vectorization opportunities as in the previous section arise.

@@

Python also provide same features for set and dictionnary.

Set comprehension can also be rewritten in the form of
generator expression using the eponym function: \texttt{\{ x + 3 for x in l\}}
is equivalent to \texttt{set(x + 3 for x in l)}.

Dictionnary comprehension and multi-level comprehension (i.e.\ containing
multiple \texttt{for} clauses) are not considered in this preliminary work.

@@

The \texttt{numpy} module played a major role in the adoption of Python by the
scientific community. In a nutshell, this is a thin Python wrapper above a raw
array that provides an array language embeded into Python. An example of
\texttt{numpy} code is show on Listing~\ref{lst:numpy-code}.

@@

Common features of array language include indexing, slicing, point-to-point
operation etc. The latter is traditionnaly a good source of vectorization.
However, because of the module layer, \texttt{numpy} always create a new array
to hold intermediate result. For instance in the expression \texttt{(1 -
x[:-1]) ** 2)} two array are created ---slicing does not create an array but a
view. The implied extra cost would hide the benefits of vectorization.

The typical way to solve this issue in C++ is to use \emph{expression
templates}, a technique described in~\cite{} that uses the C++ type system to
perform lazy evaluation of an expression. However, this approach is limited to
a single expression. Listing~\ref{lst:numpy-code-fs} illustrates the situation:
expression templates would not prevent the creation of temporary arrays
\texttt{t0} and \texttt{t1}. Fortunately, it is possible to apply \emph{forward
substitution} to fallback to the expression from Listing~\ref{lst:numpy-code}
and avoid the creattion of any temporary.

@@

It is possible to hook in the expression template to add vectorization support.
Indeed the expression from Listing~\ref{lst:numpy-code} hides a single vector
loop, that can be explicitly vectorized using Bosst.SIMD in a similar manner as
the \texttt{map} and \texttt{sum} intrinsics.

One potential drawback of the approach would be the generation of multiple load
from the same location, as in \texttt{1. / a + a}, where \texttt{a} is loaded
twice on each loop iteration. Fortunately, the backend C++ compiler is capable
of detecting the redundant load and perform \emph{redundant load store
elimination}, effectively issuing a single load.

@@

Experiments are run on a computer running a Linux kernel version 3.10. The
processors are Intel Core i7, so the AVX instruction set is available. Python
version is 2.7.5, Pythran version is the \texttt{HEAD} of the \texttt{wpmvp14}
branch of the official Pythran repository
\url{https://github.com/serge-sans-paille/pythran}. The backend C++ compiler is
Clang version 3.4-1. The compilation flags used are \texttt{-O2} without
vectorisation support and \texttt{-O2 -march=native -DUSE\_BOOST\_SIMD} with
vectorisation support. Using \texttt{-O3} or \texttt{-Ofast} actually degraded
performances.

@@

This experiment considers the code from Listing~\ref{lst:euler06-py}. This code
uses \emph{generator expression} that have not been explored in this paper, so
it has been modified to use \emph{list comprehension} instead. As a result, a
lot of temporay array are used, which leads to the performance show on
Figure~\ref{fig:euler06-timings}. Although Pythran does a decent job at
optimizing the code, vectorization does not provide significant speedup. This
is due to the poor computation over memory transfer ratio, and the lack of
optimization done by Pythran on that aspect for \texttt{list}s. A similar setup
is evaluated in the context of \texttt{numpy} arrays in
\S~\ref{sec:bench-numpy} that use temporary array elimination.

@@

\subsection{Vectorization of the \texttt{map} intrisinc}

To evaluate the vectorization of the \texttt{map} intrisinc, two expressions
are used: \texttt{map(abs, r)} and \texttt{map(lambda x: 5 * x + 3, r)}, where
\texttt{r} is a \texttt{ndarray} of random double-precision floating point
numbers. Figure~\ref{fig:map-timings} summarizes the result. The execution time
for the Cython version is extremly long due to the ndarray indexing and
function call overheads in CPython. Thanks to static typing and aggressive
inlining, Pythran does not suffer from these issues, as shown by its sequential
execution time. Vectorization is beneficial and yields a $\times2.93$ speedup.

@@

\subsection{Vectorization of the \texttt{sum} intrisinc}

To evaluate the vectorization of the \texttt{sum} intrisinc, two expressions
are used: \texttt{sum(r)} and \texttt{sum(r * r)}, where \texttt{r} is a
\texttt{ndarray} of random double-precision floating point numbers.
Figure~\ref{fig:sum-timings} summarizes the result. Again, CPython is extremely
slow. It is interesting to note that the extra array multiplication does not
impacts the excution time much, simply because it is an operation pefromed at
the native level by the \texttt{numpy} module, while the \texttt{sum} is fully
interpreted. The vectorized reductions yields a decent $\times2.41$ speedup
over the sequential version.

@@

Future works include the case of Python generators, that have the benfit of not
creating intermediate container to hold the data, but require more analyse in
the context of vectorization. Handling \texttt{list} operations in an efficient
manner as done for \texttt{numpy} arrays is also a required developpment axis.

@@

Several students from Télécom Bretagne have been working on this project, the
authors would like to thank Alan Raynaud, Adrien Merlini and Yuanchang Peng for
their contribution to the Pytrhan project. They are also in debt to Mehdi
Aminin for his careful code reviews.

>>>>

The Python language~\cite{rossum97} has grown in audience for the past ten
years, even reaching the world of scientific computations~\cite{oliphant2007}
thanks to the \texttt{numpy} module~\cite{numpyarray2011}, a module that
provides an efficient multi-dimensional array type, and the \texttt{scipy}
package~\cite{scipy} that provides a MATLAB-like API. As a consequence, more
and more code is being written either in pure Python, generally to prototype an
application, or as a Python and native code mix when performance matters.
However, Python trades performance for dynamicity and does not particularly
shines in terms of performance.

@@

    \item[dynamic binding] each variable access is made through a dictionary
        look-up to bind variable names to variable instances~;

@@

Single Instruction Multiple Data (SIMD) extensions have been a feature of
choice for processor manufacturers for a couple of decades. However,
programming applications that take advantage of the SIMD extension available on
the current target remains a complex task. Programmers that use low-level
intrinsics have to deal with a verbose programming style due to the fact that
SIMD instructions sets cover a few common functionalities, requiring to bury
the initial algorithms in architecture specific implementation details.
Furthermore, these efforts have to be repeated for every different extension
that one may want to support, making design and maintenance of such
applications very time consuming.

\boostsimd{} is a high-level C++ library to program SIMD architectures. 
Designed as an \textit{Embedded Domain Specific Language}, \boostsimd{} 
provides both expressiveness and performance by using generic programming 
to handle vectorization. \boostsimd{} does not only provide a portable 
way but also enable the use of common programming idioms when designing 
SIMD-aware algorithms. Two components are provided:

@@

Previous examples of integration within standard algorithm are still limited.
Applying \texttt{transform} or \texttt{fold} algorithm on SIMD aware data
requires the size of the data to be an exact multiple of the SIMD register width
and, in the case of fold, to potentially perform additional operations at the
end of its call. To alleviate this limitation, \boostsimd{}
provides its own overload for both \texttt{transform} and \texttt{fold} that take
care of potential trailing data and performs proper completion of \texttt{fold}.
Moreover, as \boostsimd{} provides default implementation of all its function for scalar values,
we can rely on polymorphic callable object to write a mixed scalar/SIMD call to algorithm
like \texttt{transform}. By using the \texttt{boost::simd::transform} algorithm, code
handling alignment, scalar prologue and epilogue is generated.

@@

The first attempt to have Pythran generate vectorized code directly relied on
SSE intrinsics. This approach quickly proved to be inappropriate for several
reasons:

@@

    \item[completness] Many basic mathematic functions such as cos or sin
        are not directly available in the target instruction sets;

@@

Pythran uses function objects to represent any Python functions, either
user-functions or functions from the standard library. The former are
automatically generated from the user code analysis, and the latter are
hand-coded in a library called \texttt{pythonic++} provided with the compiler.
For instance the Python function \texttt{def foo(x): return 3 * x} is compiled
into a code similar to the one in Listing~\ref{lst:simple-pythran-output}. The
actual would be more complicated to handle type inference across several
statements, but the structure remains valid.

@@

A high-level language such as Python exhibits another level of vectorization
opportunity in implicit loop. Implicit loops can be found in many places, but
this paper focuses on \emph{list comprehension} and \emph{generator
expression}, \emph{map} or \emph{sum} intrinsics and \emph{array operations}
through the \texttt{numpy} module.

@@

The \texttt{sum} intrinsic is a typical case of potential vectorization of an
implicit loop. Its prototype is: \texttt{sum(sequence[, start]) $\rightarrow$
value} and its behavior is equivalent to the Python code in
Listing~\ref{lst:sum-py}. Depending on the type of \texttt{sequence}, and also
whether one needs to strictly conforms to the IEEE Standard for Floating-Point
Arithmetic (IEEE 754), vectorization of this reduction is a comm case of \texttt{fold}.

@@

Apart from standard compliance, to be able to efficiently vectorize this code in C++, one must check that:

@@

Condition~\ref{enu:same-types} is a prerequiste for static typing and efficient
unboxing. For instance, any code that does not match it will fail to compile
using Pythran or Shed-Skin. It is also satisfied for the \texttt{ndarray}
container as they wraps unboxed arrays. Condition~\ref{enu:scalar-types} can be
checked at compile time. Python supports two native scalar types candidates for
vectorization: \texttt{int} (64 bits integer) and \texttt{float}
(double-precision floating point numbers). The \texttt{numpy} model add supports for a
wide range of signed and unsigned integers as well as single, double or
quadruple precision float. Depending on the target architecture, only some of
them can fit into vector registers. Finally condition~\ref{enu:contiguous} also
is a static property of a container. It is verified for C++ alternative to
Python's \texttt{list}, \texttt{str} and \texttt{ndarrays}, but generally not
for \texttt{set} or \texttt{dict}. Condition \ref{enu:random-access} is not
matched by \emph{generator}s, the Python incarnation of co-routines.

@@

As function have to be generic but some containers do not match required
properties, it is possible to select a vectorized
implementation of fall-back to a generic one using Substitution Failure
Is Not An Error (SFINAE)~\cite{metaprogramming2002}.

@@

Python provides \emph{list comprehension} as a way to create \texttt{list} from
any iterable. For instance \texttt{[abs(x) for x in l]} creates a new
\texttt{list} that holds the absolute value of each element of \texttt{l}. A
filter can be set up as in  \texttt{[abs(x) for x in l if x \% 3]}, where only
the elements non multiple of 3 are taken into account. Let us call
\emph{comprehension rule} the expression on the left of the
\texttt{for}.

List comprehension can be translated into a combination of \texttt{map} and
(optionally) \texttt{filter}, by transformation of the comprehension rule into
a lambda function. For instance the previous example can be re-written as
\texttt{map(abs, filter(lambda x: x\% 3, l))}. Once in this form, the same
    vectorization opportunities as in the previous section arise.

@@

Python also provide same features for set and dictionary.  Set comprehension
can be rewritten in the form of generator expression using the eponym function:
\texttt{\{ x + 3 for x in l\}} is equivalent to \texttt{set(x + 3 for x in l)}.
Dictionary comprehension and multi-level comprehension (i.e.\ containing
multiple \texttt{for} clauses) are not considered in this preliminary work.

@@

The \texttt{numpy} module played a major role in the adoption of Python by the
scientific community. In a nutshell, this is a thin Python wrapper above a raw
array that provides an array language embedded into Python. An example of
\texttt{numpy} code is show on Listing~\ref{lst:numpy-code}.

@@

Common features of array language include indexing, slicing, point-to-point
operation etc. The latter is traditionally a good source of vectorization.
However, because of the module layer, \texttt{numpy} always create a new array
to hold intermediate result. For instance in the expression \texttt{(1 -
x[:-1]) ** 2)} two array are created ---slicing does not create an array but a
view. The implied extra cost would hide the benefits of vectorization.

The typical way to solve this issue in C++ is to use \emph{expression
templates}, a technique described in~\cite{} that uses the C++ type system to
perform lazy evaluation of an expression. However, this approach is limited to
a single expression. Listing~\ref{lst:numpy-code-fs} illustrates the situation:
expression templates would not prevent the creation of temporary arrays
\texttt{t0} and \texttt{t1}. Fortunately, it is possible to apply \emph{forward
substitution} to fall-back to the expression from Listing~\ref{lst:numpy-code}
and avoid the creation of any temporary.

@@

It is possible to hook in the expression template to add vectorization support.
Indeed the expression from Listing~\ref{lst:numpy-code} hides a single vector
loop, that can be explicitly vectorized using \boostsimd in a similar manner as
the \texttt{map} and \texttt{sum} intrinsics.

One potential drawback of the approach would be the generation of multiple load
from the same location, as in \texttt{1. / a + a}, where \texttt{a} is loaded
twice on each loop iteration. Fortunately, the back-end C++ compiler is capable
of detecting the redundant load and perform \emph{redundant load store
elimination}, effectively issuing a single load.

@@

Experiments are run on a computer running a Linux kernel version 3.10. The
processors are Intel Core i7, so the AVX instruction set is available. Python
version is 2.7.5, Pythran version is the \texttt{HEAD} of the \texttt{wpmvp14}
branch of the official Pythran repository
\url{https://github.com/serge-sans-paille/pythran}. The back-end C++ compiler is
Clang version 3.4-1. The compilation flags used are \texttt{-O2} without
vectorisation support and \texttt{-O2 -march=native -DUSE\_BOOST\_SIMD} with
vectorisation support. Using \texttt{-O3} or \texttt{-Ofast} actually degraded
performances.

@@

This experiment considers the code from Listing~\ref{lst:euler06-py}. This code
uses \emph{generator expression} that have not been explored in this paper, so
it has been modified to use \emph{list comprehension} instead. As a result, a
lot of temporary array are used, which leads to the performance show on
Figure~\ref{fig:euler06-timings}. Although Pythran does a decent job at
optimizing the code, vectorization does not provide significant speedup. This
is due to the poor computation over memory transfer ratio, and the lack of
optimization done by Pythran on that aspect for \texttt{list}s. A similar setup
is evaluated in the context of \texttt{numpy} arrays in
\S~\ref{sec:bench-numpy} that use temporary array elimination.

@@

\subsection{Vectorization of the \texttt{map} intrinsic}

To evaluate the vectorization of the \texttt{map} intrinsic, two expressions
are used: \texttt{map(abs, r)} and \texttt{map(lambda x: 5 * x + 3, r)}, where
\texttt{r} is a \texttt{ndarray} of random double-precision floating point
numbers. Figure~\ref{fig:map-timings} summarizes the result. The execution time
for the CPython version is extremely long due to the \texttt{ndarray} indexing and
function call overheads in CPython. Thanks to static typing and aggressive
inlining, Pythran does not suffer from these issues, as shown by its sequential
execution time. Vectorization is beneficial and yields a $\times2.93$ speedup.

@@

\subsection{Vectorization of the \texttt{sum} intrinsic}

To evaluate the vectorization of the \texttt{sum} intrinsic, two expressions
are used: \texttt{sum(r)} and \texttt{sum(r * r)}, where \texttt{r} is a
\texttt{ndarray} of random double-precision floating point numbers.
Figure~\ref{fig:sum-timings} summarizes the result. Again, CPython is extremely
slow. It is interesting to note that the extra array multiplication does not
impacts the execution time much, simply because it is an operation performed at
the native level by the \texttt{numpy} module, while the \texttt{sum} is fully
interpreted. The vectorized reductions yields a decent $\times2.41$ speedup
over the sequential version.

@@

Future works include the case of Python generators, that have the benefit of not
creating intermediate container to hold the data, but require more analyse in
the context of vectorization. Handling \texttt{list} operations in an efficient
manner as done for \texttt{numpy} arrays is also a required development axis.

@@

Several students from T{\'e}l{\'e}com Bretagne have been working on this project, the
authors would like to thank Alan Raynaud, Adrien Merlini and Yuanchang Peng for
their contribution to the Pythran project. They are also in debt to Mehdi
Amini for his careful code reviews.
